---
title: "class11"
Author: "Yufei Liu (A16222438)"
format: html
---

# Section 1. Proportion of G/G in a population


Downloaded CSV file from Ensemble:
https://www.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=105535077#373531_tablePanel

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```
```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl)
```

# Setion4 Population Analysis

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3 expression.

How many samples do we have?
```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)

nrow(expr)
table(expr$geno)
```

make a summary boxplot for display
```{r}
library(ggplot2)

ggplot(expr, aes(geno, exp, fill = geno)) + geom_boxplot(notch = T)
```

It is clear that G allele reduces the expression level!